---
title: 'Data 607: Final Project'
author: "Anthony Munoz, Christina Valore, David Apolinar"
date: "5/10/2019"
output: 
  html_document:
    #df_print: paged
    toc: true # table of content true
    toc_collapsed : false
    toc_float: true
    code_folding : show
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
    smooth_scroll: true
    theme: cerulean  # many options for theme, this one is my favorite.
    highlight: textmate  # specifies the syntax highlighting style
---

```{r setup, include=FALSE}
library(mongolite)
library(magrittr)
library(dplyr)
library(tidyr)
```

# Project Overview

NYC Subway delays have been constantly scrutinized for being delayed all the time. There are daily articles published criticizing the MTA for not doing enough to resolve the constant delays that plague the 100+ year aging subway system. To make matters worse, there is constant scrutiny over the MTA overpaying its workers and not contributing to repairing its infrastructure.

* https://patch.com/new-york/new-york-city/nyc-subway-delays-8-lines-messed-during-morning-rush
* https://www.cbsnews.com/news/new-york-struggling-transit-agency-mta-paid-one-worker-344k-in-overtime/

As part of this analysis, we want to determine what could be the causes of the NYC Subway system. We gathered data from several locations, including the MTA's own published performance metrics.

# Business Question


As part of this analysis, we want to answer the following questions?

* Is the time of the year predicitive of the NYC subway on-time perfromance (OTP)?
* Are there other factors contributing to the delays?
  
  * population increases
  * weather impact

Using some of the collected data, we want to detemrine whether there may be external factors that may show some level of correlation with they delays. It's quite possible that it may not be the case, but this is why we analyze the data to come to a conclusion.

# Obtain the Data

Source: MTA

```{r MTA}
# load data
data<- read.csv("https://raw.githubusercontent.com/ChristinaValore/stats-prob-606/master/Performance_NYCT.csv", header=TRUE, check.names = FALSE)
# subsetting the data to only pull out the OTP (ON TIME %) for individual subway lines
otp <-data[ which(data$PARENT_SEQ=='391690'), ]
# separate the indicator and name column to have the subway names listed individually

otp<-separate(otp,INDICATOR_NAME,into=c( "INDICATOR" , "NAME" ),sep="-")
otp$NAME<-as.factor(otp$NAME)
# remove columns that are not needed for this analysis
otp<- otp[c(-1:-4)]
```

Source: https://www.weather.gov/media/okx/Climate/CentralPark/monthlyannualtemp.pdf

```{r 2}
file.old <- "/Users/davidapolinar/Dropbox/CUNYProjects/Srping2019/Data607/FinalProject/monthlyannualtemp-converted.csv"
file <- "https://raw.githubusercontent.com/dapolloxp/607-finalproject/master/monthlyannualtemp-converted.csv"

weather.data = read.csv(file, stringsAsFactors = FALSE)
cleaned.weather <- weather.data %>% select(-11)
missing.days <- c(77.5, 76, 68.5, 57.5, 47.5, 38)
cleaned.weather[150,8:13] <- missing.days
cleaned.weather[150,14] <- apply(cleaned.weather[10,2:13],1,mean)

mongo_pwd_file <- "/Users/davidapolinar/Dropbox/CUNYProjects/Srping2019/Data607/Week 13/mongopassword.txt"

mongo_password <-read.delim(mongo_pwd_file, header = FALSE, stringsAsFactors = FALSE)

url <- paste0("mongodb://nyc.admin:", mongo_password$V1 , "@10.20.1.6:27017/weather",sep = "")

mgo <- mongo(collection = "nyc", db = "weather",  url=url, verbose = TRUE)
mgo$drop()
mgo$insert(cleaned.weather)

test<-mgo$find('{"YEAR":{"$gt":2007}}')

mgo$find('{"YEAR":{}}')

```


```{r}
test2 <- test %>% select(-ANNUAL)
new <- gather(test2, 'JAN', 'FEB', 'MAR', 'APR','MAY','JUN','JUL','AUG', 'SEP','OCT','NOV','DEC',key = 'Month', value = 'Avg Monthly Temp')

new %>% arrange(YEAR)
```




```{r population}
library(rvest)
population.data <- read_html("http://worldpopulationreview.com/us-cities/new-york-city-population")
pop.table <-population.data %>% html_nodes("table") %>% .[[1]] %>% html_table()
expected.growth <-pop.table[2:9,1:2]
names(expected.growth) <- pop.table[1,1:2]

expected.growth

```

# Cleansing

```{r}
## Once we have all the data - we can attempt to combine all a do a futher cleansing process as needed.
```


# Exploration

```{r ggplot}
library(ggplot2)
summary(otp$MONTHLY_ACTUAL)
summary(otp$MONTHLY_TARGET)
# qualitative value - no need for a summary 
summary(otp$PERIOD_MONTH)
#aggregating the monthly on-time values by subway line 
month_mean<-aggregate(otp$MONTHLY_ACTUAL, by = list(otp$NAME), FUN=mean)
names(month_mean)<- c("LINE","AVG")
#Plotting the aggregate values to have a quick view of which subways have the best/worst time overall monthly
ggplot(month_mean,aes(x=reorder(LINE,-AVG), y= AVG)) + geom_bar(stat="identity") + theme (axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

# Modeling 

# Interpreration

Is weather a predictor of the NYC subway's on time percentage (OTP)?

# Obtain the Data



# Cleansing

# Exploration

# Modeling 

# Interpreting

# Conclusion